---
title: "Endo Logistic Analysis"
author: "Adwoa Odoom"
date: "2026-01-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# clear console
cat("\014")

# install packages
if (!require("pacman")) install.packages("pacman")

# install additional packages
pacman::p_load(
  rio,
  here,
  tidyverse,
  ggplot2,
  dplyr,
  ggthemes,
  gtsummary,
  flextable,
  skimr,
  gt,
  forestplot,
  broom,
  writexl,
  flextable,
  ggpubr
)

# set root directory and make into object
root_dir <- "/Users/adwoaodoom/Desktop/Ad-Hoc Projects"
```

```{r}
# import dataset
endo <- rio::import(here::here(root_dir, "structured_endometriosis_data.csv"))
```

```{r}
# explore dataset
str(endo)
dim(endo) # 10k observations, 7 columns

# rename columns
endo <- endo %>%
  rename(
    diagn = `Diagnosis`,
    mens_irr    = `Menstrual_Irregularity`,
    pain_lvl    = `Chronic_Pain_Level`,
    hormone_irr = `Hormone_Level_Abnormality`,
    infert      = `Infertility`,
    age = `Age`,
    bmi = `BMI`)

#check
names(endo)

# change applicable variables to factors for comparison
endo <- endo %>%
  mutate(
    diagn = factor(diagn, levels =c(0,1), labels = c("No", "Yes")),
    mens_irr    = factor(mens_irr, levels = c(0,1), labels = c("No","Yes")),
    hormone_irr = factor(hormone_irr, levels = c(0,1), labels = c("Normal","Irregular")),
    infert      = factor(infert, levels = c(0,1), labels = c("No","Yes")))

# check if converted correctly
is.factor(endo$diagn)
is.factor(endo$mens_irr)
is.factor(endo$hormone_irr)
is.factor(endo$infert)

# use skimr to glean more insight into data
skimr::skim(endo)
```

```{r}
# visualization: boxplot for continuous variables
# age
endo %>% 
  ggplot(aes(x = diagn, y = age)) + 
  geom_boxplot() + 
  stat_compare_means(
    method = "t.test",
    label = "p.signif"
  ) +
  labs(title = "Participant Age",
       x = "Diagnosis",
       y = "Age, years") +
  theme_bw()

ggsave("Participant_Age.jpeg", width = 8, height = 6)
```

```{r}
# chronic pain level
endo %>% 
  ggplot(aes(x = diagn, y = pain_lvl)) + 
  geom_boxplot() +
  stat_compare_means(
    method = "t.test",
    label = "p.signif"
  ) +
  labs(title = "Self-Reported Pain Level",
       x = "Diagnosis",
       y = "Pain Score") + 
  theme_bw() # background color

ggsave("Self-Reported Pain.jpeg", width = 8, height = 6)
```

```{r}
# bmi
endo %>% 
  ggplot(aes(x = diagn, y=bmi)) +
  geom_boxplot() + 
  stat_compare_means(
    method = "t.test",
    label = "p.signif"
  ) +
  labs(title = "Participant Body Mass Index Distribution",
       x = "Diagnosis",
       y = "BMI (kg/m^2)") + 
  theme_bw() # background color 

ggsave("Participant Body Mass Index Distribution.jpeg", width = 8, height = 6)
```

```{r}
t.test(age ~ diagn, data = endo)

t.test(bmi ~ diagn, data = endo)

t.test(pain_lvl ~ diagn, data = endo)
# save
ggsave()

```

```{r}
# logistic regression - if age is a predictor for an endo diagnosis
age_fit <- glm(diagn ~ age, data = endo, family = binomial(link = "logit"))


# view
summary(age_fit)

# odds ratio
coef(age_fit) %>% exp()

# By itself, age is not a significant predictor in whether an individual has an endo diagnosis. The odds ratio means that each additional year decreases odds by 1%.

# model with multiple predictors
endo_mod <- glm(diagn ~ age + mens_irr + pain_lvl + hormone_irr + infert + bmi, data = endo, family = binomial(link = "logit"))

# view
summary(endo_mod)

# odds ratio
coef(endo_mod) %>% exp()

# put model in summary table
endo_table <- tbl_regression(
  endo_mod,
  exponentiate = TRUE, # show odds ratios
    label = list(
    age          ~ "Age (years)",
    mens_irr     ~ "Menstrual Irregularity",
    pain_lvl     ~ "Chronic Pain Level",
    hormone_irr  ~ "Hormone Irregularity",
    infert       ~ "Infertility",
    bmi          ~ "BMI")) %>% 
      modify_caption("Endometriosis Logistic Regression Model") %>% 
      bold_labels() %>% 
      as_flex_table() %>%
  bg(bg = "white", part = "all") %>%
  color(color = "black", part = "all")


#view table
endo_table

# save
save_as_image(endo_table, path = "endo_table.png")


# odds ratio chart
coef_endo <- coef(endo_mod)
se_endo <- summary(endo_mod)$coefficients[, "Std. Error"]

forest_data <- data.frame(
  term = names(coef_endo),
  mean = exp(coef_endo),
  lower = exp(coef_endo - 1.96 * se_endo),
  upper = exp(coef_endo + 1.96 * se_endo))

forest_data <- forest_data[forest_data$term != "(Intercept)", ]

# rename
forest_data$term <- c(
  "Age",
  "Menstrual Irregularity",
  "Pain Level",
  "Hormone Irregularity",
  "Infertility",
  "BMI"
)

# plot it
forestplot(
  labeltext = forest_data$term,
  mean = forest_data$mean,
  lower = forest_data$lower,
  upper = forest_data$upper,
  zero = 1,
  xlab = "Odds Ratio"
)

# save
ggsave("Odds Ratio.jpeg", width = 8, height = 6)

```

Based on the model output, all variables except age were significant predictors in whether an individual receives an endometriosis diagnosis. 
Having menstrual irregularity increased the odds of having an endo diagnosis by 57% compared to those without menstrual irregularities.
A one-unit increase in pain level increased the odds of having an endo diagnosis by 14%.
Having hormone irregularities increased the odds of having an endo diagnosis by 128% compared to those without hormone irregularities.
Having infertility increased the odds of having an endo diagnosis by 56% compared to those without infertility.
A one-unit increase in BMI increased the odds of having an endo diagnosis by 5%.

If an individual mentions these symptoms to their clinician, they should be screened for endometriosis as these factors are influential in having an endometriosis diagnosis.

